---
title: "PEEL-1 Raw BigWig Coverage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PEEL-1 Raw BigWig Coverage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9.5,
  fig.height = 7.2,
  dpi = 220,
  fig.bg = "white"
)

```

# Overview

This article plots raw PEEL-1 coverage for four *C. elegans* strains over the
same chromosome-I window.  The bundled fixture contains the four BigWig files
and a WS285 GTF subset for the four genes in the interval, so the workflow is
self-contained and does not refer to source BAMs or other external paths.

The fixture BigWigs were made from one-base bins of the source BAMs without
normalization.  The plotted scores remain raw depth: `geom_coverage()` neither
normalizes, smooths, thresholds, nor expands the stored intervals to one row
per base.

```{r peel1-raw-coverage, message = FALSE, warning = FALSE, fig.alt = "A ggexon plot with four separate PEEL-1 raw coverage panels for strains XZ1516, ECA2091, ECA701, and ECA2191 above one centered four-gene annotation panel over the same chromosome-I interval.", fig.cap = "Raw, unnormalized PEEL-1 coverage for four C. elegans strains over chromosome I:2,332,338-2,373,985. Every coverage track has an independent depth scale, while one shared annotation panel shows the four-gene locus."}
library(ggexon)

fixture_dir <- system.file("extdata", "peel1_coverage", package = "ggexon")
if (!nzchar(fixture_dir)) {
  fixture_dir <- file.path(
    dirname(dirname(knitr::current_input(dir = TRUE))),
    "inst", "extdata", "peel1_coverage"
  )
}
gtf <- file.path(
  fixture_dir,
  "WS285.ugt31-zeel1-peel1-nekl1.gtf"
)
strains <- c("XZ1516", "ECA2091", "ECA701", "ECA2191")

coverage_species <- SynSpecies(name = "PEEL-1 raw coverage")
for (strain in strains) {
  individual <- SynIndividual(
    annotation_file = gtf,
    annotation_format = "gtf",
    id = strain
  )
  individual <- add_annotation(
    individual,
    SynBigWigAnnotation(
      name = "coverage",
      bigwig_file = file.path(fixture_dir, paste0(strain, ".raw.bw")),
      metadata = list(signal_unit = "raw_depth")
    )
  )
  coverage_species <- add_individual(coverage_species, individual)
}

ggexon(coverage_species) +
  geom_coverage(annotation = "coverage", fill = "#4C78A8") +
  geom_exon(
    species = "XZ1516",
    chr = "I",
    subset = c(2332338L, 2373985L),
    annotation_type = "exon"
  ) +
  facet_genomics(
    ggplot2::vars(track),
    ncol = 1,
    strip.position = "left"
  ) +
  scale_panel_coverage("free_y") +
  center_panel_annotation() +
  theme_ggexon_track() +
  theme_ggexon_side_strips("left")
```

# Reading the tracks

The BigWig values are raw, nonnegative depth.  Each file owns a first-class
coverage panel, and `scale_panel_coverage("free_y")` gives every strain its own
depth scale.  A high-depth track such as ECA2091 therefore does not compress
the lower-depth tracks.

The single `geom_exon()` request uses XZ1516 as the representative gene model
for the common chromosome-I window.  It occupies one independent annotation
panel below the four coverage panels; `center_panel_annotation()` centers that
gene model within its panel without changing annotation data or any coverage
scale.  No synthetic negative annotation band is added to a coverage panel.

For reproducibility, `manifest.tsv` in the fixture directory records each
source BAM, output file, interval, bin size, normalization setting, and
checksum.  The BAM files themselves are not distributed with the package.
