---
title: "CD44 Splice Variant Alignment Demo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CD44 Splice Variant Alignment Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9.5,
  fig.height = 5.4,
  dpi = 220
)
```

# Overview

This tutorial compares selected human `CD44` and mouse `Cd44` protein-coding
splice variants in one pairwise genomic view. The example uses two annotation
tracks, one middle `geom_nuclink()` panel, and genomic DNA alignments produced
with LASTZ from windows that include the gene plus a strand-aware 20 kb
promoter-side flank and 10 kb 3-prime-side flank.

The point is not to draw every transcript. `CD44` has many Ensembl isoforms, so
the bundled dataset keeps a small representative subset: the RefSeq-backed
canonical transcript, a short isoform, an intermediate isoform, and an expanded
isoform in each species. Exons are then classified as:

- `common`: present in every selected isoform for that species.
- `variable`: absent from at least one selected isoform for that species.

The data are bundled under `inst/extdata/cd44_pairwise_ensembl116`. Coordinates
remain in their original genome assemblies: human GRCh38 `chr11` and mouse
GRCm39 `chr2`. The mouse `Cd44` gene is on the reverse strand, so the plot
below reverses the mouse panel at draw time while keeping tick labels in
original genomic coordinates.

```{r cd44-isoforms, message = FALSE, warning = FALSE}
library(ggexon)

demo_dir <- system.file("extdata", "cd44_pairwise_ensembl116", package = "ggexon")
species <- read.delim(file.path(demo_dir, "cd44_species.tsv"), check.names = FALSE)
isoforms <- read.delim(file.path(demo_dir, "cd44_selected_isoforms.tsv"), check.names = FALSE)

isoform_table <- isoforms[, c(
  "display_name", "transcript_name", "transcript_id", "exon_count",
  "selection_reason"
)]
names(isoform_table) <- c("Species", "Transcript", "Ensembl transcript", "Exons", "Reason")
knitr::kable(isoform_table)
```

# Pairwise splice-variant plot

The plotting table has one row per exon per selected isoform. The y-position is
assigned by `selection_order`, which keeps the four selected isoforms visible
instead of overplotting them on one baseline. Exon fill maps to exon-level splice
role, while link fill maps to binned LASTZ percent identity. The transcript
backbone is drawn in neutral grey with `transcript_backbone_fill`.

```{r cd44-plot, message = FALSE, warning = FALSE, fig.alt = "A ggexon pairwise CD44 splice variant plot with four human isoform rows above a nucleotide-link panel and four mouse isoform rows below it. The mouse x axis is reversed. Common exons are green, variable exons are orange, transcript backbones are grey, and nucleotide-link ribbons are colored from pale to dark blue by LASTZ identity range.", fig.cap = "Selected human CD44 and mouse Cd44 splice variants with the mouse panel reversed. Exons present in every selected isoform are green; variable splice exons are orange. LASTZ genomic DNA alignment blocks are drawn as nucleotide links and colored by percent-identity range."}
exons <- read.delim(file.path(demo_dir, "cd44_selected_exons.tsv"), check.names = FALSE)
links <- read.delim(file.path(demo_dir, "cd44_nuclinks_lastz.tsv"), check.names = FALSE)

track_levels <- c("human", "link_human_mouse", "mouse")
track_labels <- c(
  human = sprintf("Human CD44 (%s)", species$chr[species$species == "human"]),
  link_human_mouse = "",
  mouse = sprintf("Mouse Cd44 (%s)", species$chr[species$species == "mouse"])
)

exons$track <- factor(exons$track, levels = track_levels)
links$track <- factor(links$track, levels = track_levels)
exons$exon_role <- factor(exons$exon_role, levels = c("common", "variable"))
panel_xlim <- stats::setNames(
  lapply(seq_len(nrow(species)), function(i) c(species$window_start[[i]], species$window_end[[i]])),
  species$species
)
panel_xlim_chr <- stats::setNames(species$chr, species$species)
identity_levels <- c("50-55%", "55-60%", "60-65%", "65-70%", ">=70%")
links$identity_bin <- factor(links$identity_bin, levels = identity_levels)

splice_palette <- c(common = "#246B5A", variable = "#D88A3D")
identity_palette <- c(
  "50-55%" = "#D5E3F0",
  "55-60%" = "#A8C7DD",
  "60-65%" = "#6FA6C8",
  "65-70%" = "#2F7FAA",
  ">=70%" = "#0E557A"
)

ggexon() +
  geom_nuclink(
    data = links,
    mapping = aes(
      tspecies = tspecies,
      tchr = tchr,
      tstart = tstart,
      tend = tend,
      qspecies = qspecies,
      qchr = qchr,
      qstart = qstart,
      qend = qend,
      strand = strand,
      group = group,
      fill = identity_bin
    ),
    colour = NA,
    alpha = 0.38,
    inherit.aes = FALSE
  ) +
  geom_exon(
    data = exons,
    mapping = aes(
      xmin = xmin,
      xmax = xmax,
      ymin = ymin,
      transcripts = transcripts,
      strand = strand,
      track = track,
      type = type,
      fill = exon_role
    ),
    exon_height = 0.5,
    transcript_backbone_ratio = 0.08,
    transcript_arrow_ratio = 0.5,
    transcript_arrow_length = 700,
    transcript_backbone_fill = "grey82",
    transcript_backbone_colour = NA,
    colour = "grey25",
    linewidth = 0.12
  ) +
  facet_genomics(
    vars(track),
    scales = "free_x",
    ncol = 1,
    link_panel_height = 0.32,
    link_axis = "none",
    link_strip = "blank",
    annotation_axis = "bottom",
    reverse_x = "mouse",
    reverse_x_match_by = "track",
    strip.position = "left",
    labeller = ggplot2::as_labeller(track_labels),
    xlim = panel_xlim,
    xlim_chr = panel_xlim_chr
  ) +
  scale_fill_manual(
    values = c(splice_palette, identity_palette),
    breaks = c("common", "variable", identity_levels),
    drop = FALSE,
    name = "Exon role / LASTZ identity"
  ) +
  scale_x_continuous(
    labels = function(x) paste0(round(x / 1000), " kb"),
    expand = ggplot2::expansion(mult = c(0.01, 0.01))
  ) +
  labs(x = "Genomic coordinate (mouse panel reversed)", y = NULL) +
  theme_ggexon_track(
    base_size = 8,
    show_x_axis = TRUE,
    show_x_grid = TRUE,
    show_legend = TRUE
  ) +
  theme(
    panel.spacing.y = grid::unit(0.05, "lines"),
    legend.position = "bottom",
    legend.key.height = grid::unit(3, "mm"),
    legend.key.width = grid::unit(6, "mm"),
    plot.margin = margin(6, 8, 6, 6)
  ) +
  ggplot2::guides(
    fill = ggplot2::guide_legend(nrow = 2, byrow = TRUE, override.aes = list(alpha = 1))
  ) +
  theme_ggexon_side_strips("left", base_size = 7.5)
```

# What the alignment layer contributes

`geom_nuclink()` uses a table of target/query genomic intervals. Here those
intervals come from LASTZ alignments of genomic DNA windows spanning the human
and mouse gene loci plus 20 kb promoter-side and 10 kb 3-prime-side flanks. The
demo keeps blocks with alignment length at least 80 bp and identity at least
50%, so lower-identity central genomic matches remain visible instead of being
filtered out. The link rows are deliberately kept separate from the exon
annotation rows:

- annotation rows describe selected transcript structures and exon membership;
- link rows describe nucleotide-level genomic alignment blocks and identity
  ranges;
- `facet_genomics()` creates the middle panel and injects the source-panel
  coordinates that `geom_nuclink()` needs.

```{r cd44-link-summary, message = FALSE, warning = FALSE}
homology_candidates <- read.delim(
  file.path(demo_dir, "cd44_exon_homology_ranked.tsv"),
  check.names = FALSE
)

summary_table <- data.frame(
  Metric = c(
    "Selected human isoforms",
    "Selected mouse isoforms",
    "LASTZ link blocks retained",
    "Ranked exon-pair candidates",
    "Reciprocal-best exon-pair candidates"
  ),
  Value = c(
    length(unique(isoforms$transcript_id[isoforms$species == "human"])),
    length(unique(isoforms$transcript_id[isoforms$species == "mouse"])),
    nrow(links),
    nrow(homology_candidates),
    sum(homology_candidates$reciprocal_best)
  )
)
knitr::kable(summary_table)
```

The identity bins in the plot show each retained LASTZ block. To summarize
conservation across broad gene parts, the table below clips each retained block
to three human-side regions and computes a target-bp-weighted identity. The
middle region is the longest consecutive run of variable selected exons.

```{r cd44-regional-identity, message = FALSE, warning = FALSE}
unique_exons <- read.delim(
  file.path(demo_dir, "cd44_selected_unique_exons.tsv"),
  check.names = FALSE
)

human_exons <- unique_exons[unique_exons$species == "human", ]
human_exons <- human_exons[order(human_exons$start, human_exons$end), ]
run_id <- cumsum(c(
  TRUE,
  human_exons$exon_role[-1L] != human_exons$exon_role[-nrow(human_exons)]
))
variable_runs <- do.call(rbind, lapply(split(human_exons, run_id), function(run) {
  data.frame(
    role = run$exon_role[[1L]],
    start = min(run$start),
    end = max(run$end),
    exon_count = nrow(run),
    width = sum(run$end - run$start + 1L)
  )
}))
variable_runs <- variable_runs[variable_runs$role == "variable", ]
middle <- variable_runs[order(-variable_runs$exon_count, -variable_runs$width), ][1L, ]
human_gene <- species[species$species == "human", ]
regions <- data.frame(
  Region = c("5' side", "variable exon-rich middle", "3' side"),
  start = c(human_gene$gene_start, middle$start, middle$end + 1L),
  end = c(middle$start - 1L, middle$end, human_gene$gene_end),
  check.names = FALSE
)

region_summary <- do.call(rbind, lapply(seq_len(nrow(regions)), function(i) {
  region <- regions[i, ]
  overlap_bp <- pmax(
    0L,
    pmin(links$tend, region$end) - pmax(links$tstart, region$start) + 1L
  )
  keep <- overlap_bp > 0L
  data.frame(
    Region = region$Region,
    `Retained blocks` = sum(keep),
    `Human bp covered` = sum(overlap_bp[keep]),
    `Weighted identity (%)` = round(
      stats::weighted.mean(links$identity[keep], overlap_bp[keep]),
      1
    ),
    check.names = FALSE
  )
}))
knitr::kable(region_summary)
```

The same exon-pair candidates can be summarized at the protein level. Peptide
fragments are reconstructed from selected transcript CDS codons and assigned to
the exon containing each codon start. Protein identity below uses a global
peptide alignment; the denominator is aligned amino-acid columns, including
gap columns.

```{r cd44-exon-protein-identity, message = FALSE, warning = FALSE}
protein_identity <- read.delim(
  file.path(demo_dir, "cd44_exon_protein_identity.tsv"),
  check.names = FALSE
)

protein_identity_table <- protein_identity[
  protein_identity$reciprocal_best,
  c(
    "human_exon_index", "mouse_exon_index",
    "human_exon_role", "mouse_exon_role",
    "max_link_identity",
    "human_peptide_length_aa", "mouse_peptide_length_aa",
    "protein_identical_aa", "protein_aligned_aa", "protein_gap_aa",
    "protein_identity"
  )
]
names(protein_identity_table) <- c(
  "Human exon", "Mouse exon",
  "Human role", "Mouse role",
  "Genomic identity (%)",
  "Human peptide aa", "Mouse peptide aa",
  "Identical aa", "Aligned aa", "Gap aa",
  "Protein identity (%)"
)
knitr::kable(protein_identity_table, row.names = FALSE)
```

# Reusing the pattern

For splice-variant comparisons, keep the biological feature colors on the exon
rectangles and use neutral styling for schematic structure. In `geom_exon()`,
that means setting `transcript_backbone_fill` and
`transcript_backbone_colour`. The same idea is available for
`geom_exon2()` through its intron and terminal-arrow controls, and for
`geom_genetag()` through `tag_arrow_fill` and `tag_arrow_colour`.

The reproducible data-preparation script is stored in
`data-raw/cd44_pairwise_ensembl116/build-cd44-demo.R`. It records the Ensembl
release, source URLs, strand-aware flank sizes, UCSC sequence windows, LASTZ
command, selected isoforms, common-exon calls, and exon-homology candidate
tables.
