---
title: "CD44 Splice Variant Alignment Demo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CD44 Splice Variant Alignment Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9.5,
  fig.height = 5.4,
  dpi = 220
)
```

# Overview

This tutorial compares selected human `CD44` and mouse `Cd44` protein-coding
splice variants in one pairwise genomic view. The example uses two annotation
tracks, one middle `geom_nuclink()` panel, and genomic DNA alignments produced
with LASTZ from the same windows used for plotting.

The point is not to draw every transcript. `CD44` has many Ensembl isoforms, so
the bundled dataset keeps a small representative subset: the RefSeq-backed
canonical transcript, a short isoform, an intermediate isoform, and an expanded
isoform in each species. Exons are then classified as:

- `common`: present in every selected isoform for that species.
- `variable`: absent from at least one selected isoform for that species.

The data are bundled under `inst/extdata/cd44_pairwise_ensembl116`. Coordinates
remain in their original genome assemblies: human GRCh38 `chr11` and mouse
GRCm39 `chr2`. The mouse `Cd44` gene is on the reverse strand, so the plot
below reverses the mouse panel at draw time while keeping tick labels in
original genomic coordinates.

```{r cd44-isoforms, message = FALSE, warning = FALSE}
library(ggexon)

demo_dir <- system.file("extdata", "cd44_pairwise_ensembl116", package = "ggexon")
species <- read.delim(file.path(demo_dir, "cd44_species.tsv"), check.names = FALSE)
isoforms <- read.delim(file.path(demo_dir, "cd44_selected_isoforms.tsv"), check.names = FALSE)

isoform_table <- isoforms[, c(
  "display_name", "transcript_name", "transcript_id", "exon_count",
  "selection_reason"
)]
names(isoform_table) <- c("Species", "Transcript", "Ensembl transcript", "Exons", "Reason")
knitr::kable(isoform_table)
```

# Pairwise splice-variant plot

The plotting table has one row per exon per selected isoform. The y-position is
assigned by `selection_order`, which keeps the four selected isoforms visible
instead of overplotting them on one baseline. The fill maps to the exon-level
splice role, while the transcript backbone is drawn in neutral grey with
`transcript_backbone_fill`.

```{r cd44-plot, message = FALSE, warning = FALSE, fig.alt = "A ggexon pairwise CD44 splice variant plot with four human isoform rows above a nucleotide-link panel and four mouse isoform rows below it. The mouse x axis is reversed. Common exons are green, variable exons are orange, transcript backbones are grey, and blue ribbons connect LASTZ genomic alignment blocks.", fig.cap = "Selected human CD44 and mouse Cd44 splice variants with the mouse panel reversed. Exons present in every selected isoform are green; variable splice exons are orange. LASTZ genomic DNA alignment blocks are drawn as nucleotide links in the middle panel."}
exons <- read.delim(file.path(demo_dir, "cd44_selected_exons.tsv"), check.names = FALSE)
links <- read.delim(file.path(demo_dir, "cd44_nuclinks_lastz.tsv"), check.names = FALSE)

track_levels <- c("human", "link_human_mouse", "mouse")
track_labels <- c(
  human = sprintf("Human CD44 (%s)", species$chr[species$species == "human"]),
  link_human_mouse = "",
  mouse = sprintf("Mouse Cd44 (%s)", species$chr[species$species == "mouse"])
)

exons$track <- factor(exons$track, levels = track_levels)
links$track <- factor(links$track, levels = track_levels)
exons$exon_role <- factor(exons$exon_role, levels = c("common", "variable"))

splice_palette <- c(common = "#246B5A", variable = "#D88A3D")

ggexon() +
  geom_nuclink(
    data = links,
    mapping = aes(
      tspecies = tspecies,
      tchr = tchr,
      tstart = tstart,
      tend = tend,
      qspecies = qspecies,
      qchr = qchr,
      qstart = qstart,
      qend = qend,
      strand = strand,
      group = group
    ),
    fill = "#6A8FB8",
    colour = NA,
    alpha = 0.32,
    inherit.aes = FALSE,
    show.legend = FALSE
  ) +
  geom_exon(
    data = exons,
    mapping = aes(
      xmin = xmin,
      xmax = xmax,
      ymin = ymin,
      transcripts = transcripts,
      strand = strand,
      track = track,
      type = type,
      fill = exon_role
    ),
    exon_height = 0.5,
    transcript_backbone_ratio = 0.08,
    transcript_backbone_fill = "grey82",
    transcript_backbone_colour = NA,
    colour = "grey25",
    linewidth = 0.12
  ) +
  facet_genomics(
    vars(track),
    scales = "free_x",
    ncol = 1,
    link_panel_height = 0.32,
    link_axis = "none",
    link_strip = "blank",
    annotation_axis = "bottom",
    reverse_x = "mouse",
    reverse_x_match_by = "track",
    strip.position = "left",
    labeller = ggplot2::as_labeller(track_labels)
  ) +
  scale_fill_manual(values = splice_palette, drop = FALSE, name = "Selected exons") +
  scale_x_continuous(
    labels = function(x) paste0(round(x / 1000), " kb"),
    expand = ggplot2::expansion(mult = c(0.01, 0.01))
  ) +
  labs(x = "Genomic coordinate (mouse panel reversed)", y = NULL) +
  theme_ggexon_track(
    base_size = 8,
    show_x_axis = TRUE,
    show_x_grid = TRUE,
    show_legend = TRUE
  ) +
  theme(
    panel.spacing.y = grid::unit(0.05, "lines"),
    legend.position = "bottom",
    legend.key.height = grid::unit(3, "mm"),
    legend.key.width = grid::unit(6, "mm"),
    plot.margin = margin(6, 8, 6, 6)
  ) +
  theme_ggexon_side_strips("left", base_size = 7.5)
```

# What the alignment layer contributes

`geom_nuclink()` uses a table of target/query genomic intervals. Here those
intervals come from LASTZ alignments of genomic DNA windows spanning the human
and mouse gene loci. The link rows are deliberately kept separate from the exon
annotation rows:

- annotation rows describe selected transcript structures and exon membership;
- link rows describe nucleotide-level genomic alignment blocks;
- `facet_genomics()` creates the middle panel and injects the source-panel
  coordinates that `geom_nuclink()` needs.

```{r cd44-link-summary, message = FALSE, warning = FALSE}
homology_candidates <- read.delim(
  file.path(demo_dir, "cd44_exon_homology_ranked.tsv"),
  check.names = FALSE
)

summary_table <- data.frame(
  Metric = c(
    "Selected human isoforms",
    "Selected mouse isoforms",
    "LASTZ link blocks retained",
    "Ranked exon-pair candidates",
    "Reciprocal-best exon-pair candidates"
  ),
  Value = c(
    length(unique(isoforms$transcript_id[isoforms$species == "human"])),
    length(unique(isoforms$transcript_id[isoforms$species == "mouse"])),
    nrow(links),
    nrow(homology_candidates),
    sum(homology_candidates$reciprocal_best)
  )
)
knitr::kable(summary_table)
```

# Reusing the pattern

For splice-variant comparisons, keep the biological feature colors on the exon
rectangles and use neutral styling for schematic structure. In `geom_exon()`,
that means setting `transcript_backbone_fill` and
`transcript_backbone_colour`. The same idea is available for
`geom_exon2()` through its intron and terminal-arrow controls, and for
`geom_genetag()` through `tag_arrow_fill` and `tag_arrow_colour`.

The reproducible data-preparation script is stored in
`data-raw/cd44_pairwise_ensembl116/build-cd44-demo.R`. It records the Ensembl
release, source URLs, UCSC sequence windows, LASTZ command, selected isoforms,
common-exon calls, and exon-homology candidate tables.
