---
title: "Yeast cox1 Intron Architecture Demo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Yeast cox1 Intron Architecture Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 4.2,
  dpi = 220
)
```

# Overview

This tutorial compares mitochondrial `cox1` in fission yeast
(*Schizosaccharomyces pombe*) with mitochondrial `COX1` in budding yeast
(*Saccharomyces cerevisiae* S288C). It is a compact counterpart to the CD44
pairwise alignment tutorial, but the biological contrast is different: this is
not an alternative-splicing example. It shows conserved cytochrome c oxidase
subunit 1 coding sequence across different organellar intron architectures.

The bundled dataset keeps one annotated protein-coding transcript per species.
PomBase annotates fission yeast `cox1` with three CDS blocks and two introns;
SGD annotates budding yeast S288C `COX1` with eight CDS blocks and seven
introns. The data are bundled under `inst/extdata/cox1_yeast_pairwise`.

```{r cox1-genes, message = FALSE, warning = FALSE}
library(ggexon)

demo_dir <- system.file("extdata", "cox1_yeast_pairwise", package = "ggexon")
species <- read.delim(file.path(demo_dir, "cox1_species.tsv"), check.names = FALSE)
transcripts <- read.delim(file.path(demo_dir, "cox1_transcripts.tsv"), check.names = FALSE)

gene_table <- transcripts[, c(
  "display_name", "scientific_name", "source_database", "gene_symbol",
  "gene_id", "exon_count", "intron_count"
)]
names(gene_table) <- c("Species", "Scientific name", "Source", "Gene", "Gene ID", "CDS blocks", "Introns")
knitr::kable(gene_table)
```

# Pairwise mitochondrial plot

The plotting table has one row per CDS interval. The introns are visible as gaps
on the transcript backbone. LASTZ links are computed from genomic mitochondrial
DNA windows with 500 bp flanks around each gene.

```{r cox1-plot, message = FALSE, warning = FALSE, fig.alt = "A ggexon pairwise cox1 plot with a fission yeast mitochondrial cox1 track above a nucleotide-link panel and a budding yeast mitochondrial COX1 track below it. Fission yeast has three CDS blocks, budding yeast has eight CDS blocks, and four blue LASTZ link ribbons connect conserved coding regions.", fig.cap = "Fission and budding yeast mitochondrial cox1/COX1 gene structures. CDS intervals are green, introns are represented by backbone gaps, and LASTZ genomic DNA alignment blocks are colored by percent-identity range."}
exons <- read.delim(file.path(demo_dir, "cox1_plot_exons.tsv"), check.names = FALSE)
links <- read.delim(file.path(demo_dir, "cox1_nuclinks_lastz.tsv"), check.names = FALSE)

track_levels <- c("fission_yeast", "link_fission_budding", "budding_yeast")
track_labels <- c(
  fission_yeast = sprintf("Fission yeast cox1 (%s)", species$chr[species$species == "fission_yeast"]),
  link_fission_budding = "",
  budding_yeast = sprintf("Budding yeast COX1 (%s)", species$chr[species$species == "budding_yeast"])
)

exons$track <- factor(exons$track, levels = track_levels)
links$track <- factor(links$track, levels = track_levels)
exons$exon_role <- factor(exons$exon_role, levels = "CDS")
identity_levels <- c("50-55%", "55-60%", "60-65%", "65-70%", ">=70%")
links$identity_bin <- factor(links$identity_bin, levels = identity_levels)

feature_palette <- c(CDS = "#246B5A")
identity_palette <- c(
  "50-55%" = "#D5E3F0",
  "55-60%" = "#A8C7DD",
  "60-65%" = "#6FA6C8",
  "65-70%" = "#2F7FAA",
  ">=70%" = "#0E557A"
)

ggexon() +
  geom_nuclink(
    data = links,
    mapping = aes(
      tspecies = tspecies,
      tchr = tchr,
      tstart = tstart,
      tend = tend,
      qspecies = qspecies,
      qchr = qchr,
      qstart = qstart,
      qend = qend,
      strand = strand,
      group = group,
      fill = identity_bin
    ),
    colour = NA,
    alpha = 0.42,
    inherit.aes = FALSE
  ) +
  geom_exon(
    data = exons,
    mapping = aes(
      xmin = xmin,
      xmax = xmax,
      ymin = ymin,
      transcripts = transcripts,
      strand = strand,
      track = track,
      type = type,
      fill = exon_role
    ),
    exon_height = 0.52,
    transcript_backbone_ratio = 0.08,
    transcript_backbone_fill = "grey82",
    transcript_backbone_colour = NA,
    colour = "grey25",
    linewidth = 0.12
  ) +
  facet_genomics(
    vars(track),
    scales = "free_x",
    ncol = 1,
    link_panel_height = 0.38,
    link_axis = "none",
    link_strip = "blank",
    annotation_axis = "bottom",
    strip.position = "left",
    labeller = ggplot2::as_labeller(track_labels)
  ) +
  scale_fill_manual(
    values = c(feature_palette, identity_palette),
    breaks = c("CDS", identity_levels),
    drop = FALSE,
    name = "Feature / LASTZ identity"
  ) +
  scale_x_continuous(
    labels = function(x) paste0(round(x / 1000), " kb"),
    expand = ggplot2::expansion(mult = c(0.01, 0.01))
  ) +
  labs(x = "Mitochondrial genomic coordinate", y = NULL) +
  theme_ggexon_track(
    base_size = 8,
    show_x_axis = TRUE,
    show_x_grid = TRUE,
    show_legend = TRUE
  ) +
  theme(
    panel.spacing.y = grid::unit(0.05, "lines"),
    legend.position = "bottom",
    legend.key.height = grid::unit(3, "mm"),
    legend.key.width = grid::unit(6, "mm"),
    plot.margin = margin(6, 8, 6, 6)
  ) +
  ggplot2::guides(
    fill = ggplot2::guide_legend(nrow = 2, byrow = TRUE, override.aes = list(alpha = 1))
  ) +
  theme_ggexon_side_strips("left", base_size = 7.5)
```

# What the alignment layer contributes

The nucleotide links are deliberately separate from the annotation rows:

- annotation rows describe CDS intervals and intron gaps from PomBase and SGD;
- link rows describe LASTZ genomic DNA alignment blocks and identity ranges;
- `facet_genomics()` creates the middle panel and injects the source-panel
  coordinates used by `geom_nuclink()`.

```{r cox1-link-summary, message = FALSE, warning = FALSE}
homology_candidates <- read.delim(
  file.path(demo_dir, "cox1_exon_homology_ranked.tsv"),
  check.names = FALSE
)

summary_table <- data.frame(
  Metric = c(
    "Fission yeast CDS blocks",
    "Budding yeast CDS blocks",
    "LASTZ link blocks retained",
    "Ranked exon-pair candidates",
    "Reciprocal-best exon-pair candidates",
    "Length-weighted LASTZ identity"
  ),
  Value = c(
    transcripts$exon_count[transcripts$species == "fission_yeast"],
    transcripts$exon_count[transcripts$species == "budding_yeast"],
    nrow(links),
    nrow(homology_candidates),
    sum(homology_candidates$reciprocal_best),
    sprintf(
      "%.1f%%",
      weighted.mean(links$identity, links$alignment_length)
    )
  )
)
knitr::kable(summary_table)
```

The ranked exon-pair table is derived by intersecting retained LASTZ blocks with
the CDS intervals on both tracks.

```{r cox1-exon-pairs, message = FALSE, warning = FALSE}
pair_table <- homology_candidates[, c(
  "fission_exon_rank", "budding_exon_rank", "max_min_overlap_bp",
  "max_link_identity", "reciprocal_best"
)]
names(pair_table) <- c("Fission CDS", "Budding CDS", "Shared overlap bp", "Max identity", "Reciprocal best")
knitr::kable(pair_table)
```

# Data provenance

The dataset is generated by `data-raw/cox1_yeast_pairwise/build-cox1-yeast-demo.R`.
It records the PomBase GFF3, PomBase mitochondrial FASTA, PomBase
`pombe-cerevisiae-orthologs.tsv`, SGD feature table, SGD mitochondrial FASTA,
their retrieval dates and exact MD5 checksums, and a portable LASTZ command in
`cox1_provenance.tsv`.
