---
title: "ggexon Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggexon Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

`ggexon` is designed for synteny-aware genome plotting with support for
multiple annotation layers, derived CDS and protein sequences, and patch-based
curation workflows.

This vignette sketches the core workflow:

1. Create a `SynIndividual`
2. Load a feature annotation
3. Attach additional annotation layers
4. Set human-readable labels
5. Patch corrected gene models
6. Translate proteins from selected genes
7. Combine multiple individuals in a `SynSpecies` object

# Create a `SynIndividual`

```{r eval = FALSE}
library(ggexon)

x <- SynIndividual(
  genome_file = system.file("extdata", "XZ1516.fasta", package = "ggexon"),
  annotation_file = system.file(
    "extdata",
    "caenorhabditis_XZ1516.gff3",
    package = "ggexon"
  )
)
```

# Load the active feature annotation

```{r eval = FALSE}
x <- load_annotation(x)
ann <- get_annotation(x)
```

At this point the active layer is a `SynFeatureAnnotation`.

# Attach additional annotation layers

You can attach variant, signal, and protein-domain layers to the same
`SynIndividual`.

```{r eval = FALSE}
x <- add_annotation(
  x,
  SynVCFAnnotation(
    name = "variants",
    vcf_file = system.file(
      "extdata",
      "DL238.rename.ChrV.XZ1516.vcf.gz",
      package = "ggexon"
    )
  )
)

x <- add_annotation(
  x,
  SynProteinDomainAnnotation(
    name = "interpro",
    domain_file = system.file(
      "extdata",
      "InterProScan.tsv",
      package = "ggexon"
    ),
    keytype = "protein_id",
    source_db = "InterPro"
  )
)
```

# Set plot labels

Stable IDs should remain unchanged. Use `set_gene_labels()` to add readable
labels for plotting.

```{r eval = FALSE}
x <- set_gene_labels(
  x,
  c(
    FUN_000001 = "unc-1",
    FUN_000002 = "unc-2"
  )
)
```

The active feature annotation will then carry a `plot_label` metadata column.

# Apply a corrected patch from a small GFF

Small corrected GFF fragments can be imported and applied as structured
patches.

```{r eval = FALSE}
x <- patch_annotation_from_gff(
  x,
  patch_file = system.file("extdata", "XZ1516.TA.gff", package = "ggexon"),
  mode = "replace",
  name = "ta-correction"
)
```

This replaces matching target gene models in the active feature annotation and
clears stale sequence caches.

# Translate proteins for selected genes

```{r eval = FALSE}
x <- translate_protein(
  x,
  genes = c("FUN_000001", "FUN_000002")
)

protein_seq(get_annotation(x))
```

# Query attached non-feature layers

Region-based layers are queried through their type-specific verbs.

```{r eval = FALSE}
variant_layer <- get_annotation(x, "variants")
query_variants(variant_layer, chr = "V_RagTag", start = 21574336, end = 21574400)
```

# Group individuals in a `SynSpecies`

Once you have multiple genomes or species, `SynSpecies` can hold both the
`SynIndividual` objects and their cross-species alignments.

```{r eval = FALSE}
x2 <- SynIndividual(other_genome_file, other_annotation_file)

sp <- SynSpecies(name = "Caenorhabditis")
sp <- add_individual(sp, x)
sp <- add_individual(sp, x2)

sp <- add_pairwise_alignment(
  sp,
  SynPairAlignment(
    name = "XZ1516_vs_other",
    query_individual = "XZ1516",
    target_individual = syn_id(x2),
    file = "XZ1516_vs_other.paf"
  )
)
```

For multiple-species alignments:

```{r eval = FALSE}
sp <- add_multiple_alignment(
  sp,
  SynMultiAlignment(
    name = "worm-maf",
    individuals = c("XZ1516", "N2", "CB4856"),
    file = "worms.maf"
  )
)
```

# Notes

- `SynFeatureAnnotation` is the structural layer used for CDS extraction and
  protein translation.
- `SynVCFAnnotation` and `SynBigWigAnnotation` are designed for lazy,
  region-based access.
- `SynProteinDomainAnnotation` provides protein-space annotation tied to
  transcript or protein identifiers.
- `SynSpecies` is the comparison container that binds individuals and
  alignments together.

# Next steps

Suggested follow-up material for this vignette:

- comparing two feature annotation layers on the same genome
- plotting patched genes before and after correction
- combining synteny plots with variant and signal layers
