---
title: "ggexon Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggexon Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

`ggexon` is designed for synteny-aware genome plotting with support for
multiple annotation layers, derived CDS and protein sequences, and patch-based
curation workflows.

This vignette sketches the core workflow:

1. Create a `SynIndividual`
2. Load a feature annotation
3. Attach additional annotation layers
4. Set human-readable labels
5. Patch corrected gene models
6. Translate proteins from selected genes
7. Combine multiple individuals in a `SynSpecies` object
8. Query and filter pairwise alignments
9. Plot directly from a `SynSpecies` container

# Create a `SynIndividual`

```{r eval = FALSE}
library(ggexon)

x <- SynIndividual(
  genome_file = system.file("extdata", "XZ1516.fasta", package = "ggexon"),
  annotation_file = system.file(
    "extdata",
    "caenorhabditis_XZ1516.gff3",
    package = "ggexon"
  )
)
```

# Load the active feature annotation

```{r eval = FALSE}
x <- load_annotation(x)
ann <- get_annotation(x)
```

At this point the active layer is a `SynFeatureAnnotation`.

# Attach additional annotation layers

You can attach variant, signal, and protein-domain layers to the same
`SynIndividual`.

```{r eval = FALSE}
x <- add_annotation(
  x,
  SynVCFAnnotation(
    name = "variants",
    vcf_file = system.file(
      "extdata",
      "DL238.rename.ChrV.XZ1516.vcf.gz",
      package = "ggexon"
    )
  )
)

x <- add_annotation(
  x,
  SynProteinDomainAnnotation(
    name = "interpro",
    domain_file = system.file(
      "extdata",
      "InterProScan.tsv",
      package = "ggexon"
    ),
    keytype = "protein_id",
    source_db = "InterPro"
  )
)
```

# Set plot labels

Stable IDs should remain unchanged. Use `set_gene_labels()` to add readable
labels for plotting.

```{r eval = FALSE}
x <- set_gene_labels(
  x,
  c(
    FUN_000001 = "unc-1",
    FUN_000002 = "unc-2"
  )
)
```

The active feature annotation will then carry a `plot_label` metadata column.

# Apply a corrected patch from a small GFF

Small corrected GFF fragments can be imported and applied as structured
patches.

```{r eval = FALSE}
x <- patch_annotation_from_gff(
  x,
  patch_file = system.file("extdata", "XZ1516.TA.gff", package = "ggexon"),
  mode = "replace",
  name = "ta-correction"
)
```

This replaces matching target gene models in the active feature annotation and
clears stale sequence caches.

# Translate proteins for selected genes

```{r eval = FALSE}
x <- translate_protein(
  x,
  genes = c("FUN_000001", "FUN_000002")
)

protein_seq(get_annotation(x))
```

# Query attached non-feature layers

Region-based layers are queried through their type-specific verbs.

```{r eval = FALSE}
variant_layer <- get_annotation(x, "variants")
query_variants(variant_layer, chr = "V_RagTag", start = 21574336, end = 21574400)
```

# Group individuals in a `SynSpecies`

Once you have multiple genomes or species, `SynSpecies` can hold both the
`SynIndividual` objects and their cross-species alignments.

```{r eval = FALSE}
x2 <- SynIndividual(other_genome_file, other_annotation_file)

sp <- SynSpecies(name = "Caenorhabditis")
sp <- add_individual(sp, x)
sp <- add_individual(sp, x2)

sp <- add_pairwise_alignment(
  sp,
  SynPairAlignment(
    name = "XZ1516_vs_other",
    query_individual = "XZ1516",
    target_individual = syn_id(x2),
    file = "XZ1516_vs_other.paf"
  )
)
```

For multiple-species alignments:

```{r eval = FALSE}
sp <- add_multiple_alignment(
  sp,
  SynMultiAlignment(
    name = "worm-maf",
    individuals = c("XZ1516", "N2", "CB4856"),
    file = "worms.maf"
  )
)
```

# Query pairwise alignments

Pairwise alignments can now be read back as PAF-like tables from either a
`SynSpecies` object or a `SynPairAlignment` object.

Use `subset =` to trim on both query and target coordinates, and use
`filter =` to discard short fragments by `alen`.

```{r eval = FALSE}
pairwise_alignment_data(
  sp,
  alignment = "XZ1516_vs_other",
  subset = c(
    XZ1516 = "RagTag_V:21550000-21680000",
    other = "V:20450000-20510000"
  ),
  filter = 200
)
```

The helper wrappers `subset_pairwise_alignment()` and
`filter_pairwise_alignment()` expose the same operations separately when that
reads more clearly in user code.

# Plot directly from a `SynSpecies`

You can now pass a `SynSpecies` object directly to `ggexon()` and let a
syn-aware geom resolve the plotting data lazily at build time.

For example, this draws the exon models for `XZ1516` across the
`RagTag_V:21550000-21680000` region. Because this `SynSpecies` contains only
one individual, you do not need to specify `species =`. Internally, the
plotting code resolves that sole individual, maps the user-facing chromosome
label `RagTag_V` to the annotation seqname used in the GFF (`V_RagTag` in the
bundled example), and creates the default `aes()` mapping needed by
`geom_exon()`.

```{r eval = FALSE}
x <- SynIndividual(
  genome_file = system.file("extdata", "XZ1516.fasta", package = "ggexon"),
  annotation_file = system.file(
    "extdata",
    "caenorhabditis_XZ1516.gff3",
    package = "ggexon"
  ),
  id = "XZ1516"
)

sp <- SynSpecies(name = "Caenorhabditis")
sp <- add_individual(sp, x)

ggexon(sp) +
  geom_exon(
    chr = "RagTag_V",
    subset = c(21550000, 21680000)
  )
```

If a `SynSpecies` contains multiple individuals, then `species =` becomes
required so the geom knows which one to plot.

The same direct plotting workflow is also available for `geom_gene()`, which
collapses the region to one directional span per `gene_id`.

```{r eval = FALSE}
ggexon(sp) +
  geom_gene(
    chr = "RagTag_V",
    subset = c(21550000, 21680000)
  )
```

Use the two geoms for different levels of detail:

- `geom_exon()` keeps exon-level structure and is useful when transcript
  architecture matters.
- `geom_gene()` reduces each gene to a start-to-end span with an arrow showing
  strand direction.

Both geoms now use absolute genomic coordinates, so the x-axis stays in the
original coordinate system of the chromosome rather than being re-based to a
local zero point for each track.

For pairwise comparison, a syn-aware geom can also resolve both annotation
tracks plus the middle link panel from one comparative call. The reference
species supplies the input coordinates.

```{r eval = FALSE}
ggexon(sp) +
  geom_gene(
    species = c("N2", "XZ1516"),
    reference = "XZ1516",
    chr = "RagTag_V",
    subset = c(21574445, 21584356),
    alignment = "XZ1516_vs_N2"
  ) +
  facet_genomics(ggplot2::vars(track), scales = "free_y")
```

For custom pairwise views, the PAF helper and the plotting grammar are meant
to work together: derive or specify the windows you want, filter short PAF
rows with `filter =`, and then draw the comparison.

At the moment, this direct plotting path is implemented for `geom_exon()` and
`geom_gene()`. The same pattern can be extended to other geoms so they can
consume `SynSpecies` or `SynIndividual` objects without requiring a manual
`fortify()` step.

## Extending syn-aware geoms

For package development, the direct plotting behaviour is implemented through
the custom `LayerSyn` class in `R/syn-layer.R`. That layer does two extra jobs
for syn objects: it resolves `SynSpecies` / `SynIndividual` inputs into regular
data frames, and it injects a default `aes()` mapping when the geom is called
without one.

To make a new `geom_*()` syn-aware, use `layer_class = LayerSyn` in the public
wrapper and then add two pieces in `R/syn-layer.R`:

- a `default_syn_aesthetics()` branch describing which columns should be
  auto-mapped for the geom
- a `resolve_syn_layer_data()` branch describing how to turn a syn object into
  the data frame expected by that geom

In practice, this means the geom itself can stay focused on the usual ggplot2
responsibilities, namely `setup_data()` and `draw_panel()`, while the syn layer
handles object resolution and default mapping.

# Notes

- `SynFeatureAnnotation` is the structural layer used for CDS extraction and
  protein translation.
- `SynVCFAnnotation` and `SynBigWigAnnotation` are designed for lazy,
  region-based access.
- `SynProteinDomainAnnotation` provides protein-space annotation tied to
  transcript or protein identifiers.
- `SynSpecies` is the comparison container that binds individuals and
  alignments together.

# Next steps

Suggested follow-up material for this vignette:

- comparing two feature annotation layers on the same genome
- plotting patched genes before and after correction
- combining synteny plots with variant and signal layers
