---
title: "HOX Cluster Expansion Across Chordates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HOX Cluster Expansion Across Chordates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10.5,
  fig.height = 7.4,
  dpi = 220
)
```

# Why a cluster matrix?

This example replaces the former HOXA ribbon tutorial with a view of HOX
cluster expansion. Early vertebrate whole-genome duplications produced the four
canonical gnathostome cluster families, HOXA through HOXD, and a later teleost
duplication produced paired cluster copies followed by differential loss. The
seven retained zebrafish clusters were an early line of evidence for that
teleost duplication ([Amores et al. 1998](https://pubmed.ncbi.nlm.nih.gov/9831563/));
genome-wide conserved synteny now resolves the earlier vertebrate duplications
([Simakov et al. 2020](https://www.nature.com/articles/s41559-020-1156-z)).

The figure therefore uses columns, rather than nucleotide links, to represent
the duplication history. Human, mouse, chicken, and spotted gar have one row
each. Zebrafish has separate `a` and `b` rows for its duplicated clusters. The
invertebrate chordate *Branchiostoma lanceolatum* (amphioxus) has one ancestral
cluster. To keep the matrix compact, that cluster is displayed in the HOXA
column as a layout anchor; this placement does not assign it specifically to
HOXA. Its HOXB--HOXD cells are structural blanks.

```{r hox-data, message = FALSE, warning = FALSE}
library(ggexon)

demo_dir <- system.file("extdata", "hox_cluster_expansion", package = "ggexon")
genes <- read.delim(file.path(demo_dir, "hox_genes.tsv"), check.names = FALSE)
panels <- read.delim(
  file.path(demo_dir, "hox_clusters.tsv"),
  check.names = FALSE,
  na.strings = c("", "NA")
)
species <- read.delim(file.path(demo_dir, "hox_species.tsv"), check.names = FALSE)
annotation_gaps <- read.delim(
  file.path(demo_dir, "hox_annotation_gaps.tsv"),
  check.names = FALSE
)

row_levels <- c(
  "human", "mouse", "chicken", "gar",
  "zebrafish_a", "zebrafish_b", "amphioxus"
)
row_labels <- c(
  human = "Human",
  mouse = "Mouse",
  chicken = "Chicken",
  gar = "Spotted gar",
  zebrafish_a = "Zebrafish a",
  zebrafish_b = "Zebrafish b",
  amphioxus = "Amphioxus\n(ancestral)"
)
column_levels <- c("A", "B", "C", "D")
column_labels <- c(
  A = "HOXA",
  B = "HOXB",
  C = "HOXC",
  D = "HOXD"
)
slot_order <- paste0("Hox", 15:1)

genes$species_row <- factor(genes$species_row, levels = row_levels)
genes$cluster_column <- factor(genes$cluster_column, levels = column_levels)
genes$track <- paste(genes$species_row, genes$cluster_column, sep = "::")
genes$slot <- factor(genes$slot, levels = slot_order)
genes$y <- 1

panels$species_row <- factor(panels$matrix_row, levels = row_levels)
panels$cluster_column <- factor(panels$matrix_column, levels = column_levels)
```

# Exact shared Hox slots

`geom_genebox()` uses a true square measured in millimetres. Its internal arrow
shows transcriptional direction, and black or white is selected automatically
to contrast with the fill. The figure maps fill to Hox paralog number with an
ordered, color-vision-friendly viridis HCL palette. No large legend is needed
because the shared Hox15-to-Hox1 slot labels appear on every column axis.

`strip_scale_x(slot_order = ...)` supplies a complete synthetic comparison
template. Each displayed gene's selected coding midpoint is mapped to its
curated Hox-number slot, so genes assigned to the same slot share one x
coordinate across species and clusters. Unoccupied template slots do not close
up and do not by themselves imply gene loss. This alignment is a visualization
convention, not an inferred one-to-one orthology relationship, particularly for
posterior amphioxus Hox genes. Box position therefore represents the curated
Hox paralog slot, while the internal arrow independently represents
transcription direction. The mapped BraLan2 amphioxus genes are all on the
minus strand, so their arrows point left even though their Hox-number slots
remain aligned with the vertebrate rows. If the template reverses a cluster's
native genomic direction, the gene-box arrows are reversed with it. Duplicated
genes assigned to the same slot are stacked vertically at that x coordinate.

```{r hox-matrix, message = FALSE, warning = FALSE, fig.alt = "A seven-row by four-column matrix of chordate Hox clusters. Human, mouse, chicken, spotted gar, and two zebrafish subrows occupy HOXA through HOXD columns. The ancestral amphioxus cluster is displayed under HOXA with left-pointing strand arrows, while its homologous Hox15 through Hox1 slots remain vertically aligned with vertebrate slots. HOXB through HOXD are structural blanks for amphioxus, and zebrafish HOXDB is marked as not retained.", fig.cap = "HOX cluster expansion across six chordate species. Columns represent the duplicated HOXA--HOXD cluster families. The ancestral amphioxus cluster is displayed under HOXA for layout only, not because it is specifically orthologous to HOXA. Box position represents the curated Hox paralog slot; the internal arrow independently represents transcription direction. Aligned empty slots represent genes not present in a plotted annotation, shaded cells are structurally not applicable, and zebrafish HOXDB is marked as a cluster not retained."}
active_panels <- panels[
  panels$cell_status != "structural_blank",
  c("species_row", "cluster_column")
]
slot_guides <- merge(
  active_panels,
  data.frame(xintercept = seq_along(slot_order)),
  by = NULL,
  all = TRUE
)
structural_panels <- panels[panels$cell_status == "structural_blank", ]
lost_panels <- panels[panels$cell_status == "cluster_not_retained", ]

hox_palette <- setNames(
  grDevices::hcl.colors(length(slot_order), "viridis"),
  slot_order
)

ggexon() +
  geom_rect(
    data = structural_panels,
    xmin = -Inf,
    xmax = Inf,
    ymin = -Inf,
    ymax = Inf,
    inherit.aes = FALSE,
    fill = "#F4F4F4",
    colour = NA
  ) +
  geom_rect(
    data = lost_panels,
    xmin = -Inf,
    xmax = Inf,
    ymin = -Inf,
    ymax = Inf,
    inherit.aes = FALSE,
    fill = "#FBE9E7",
    colour = "#C95F54",
    linewidth = 0.35,
    linetype = 2
  ) +
  geom_vline(
    data = slot_guides,
    aes(xintercept = xintercept),
    inherit.aes = FALSE,
    colour = "#D9D9D9",
    linewidth = 0.2
  ) +
  geom_genebox(
    data = genes,
    aes(fill = slot),
    box_size = 2.8,
    colour = "grey20",
    linewidth = 0.25,
    show.legend = FALSE
  ) +
  geom_text(
    data = structural_panels,
    aes(x = 8, y = 1, label = "not applicable"),
    inherit.aes = FALSE,
    colour = "#6F6F6F",
    size = 2.2
  ) +
  geom_text(
    data = lost_panels,
    aes(x = 8, y = 1, label = "HOXDB not retained"),
    inherit.aes = FALSE,
    colour = "#A33F36",
    size = 2.2
  ) +
  strip_scale_x(slot_order = slot_order, guide = "none") +
  facet_grid(
    rows = vars(species_row),
    cols = vars(cluster_column),
    scales = "fixed",
    drop = FALSE,
    switch = "y",
    labeller = labeller(
      species_row = as_labeller(row_labels),
      cluster_column = as_labeller(column_labels)
    )
  ) +
  scale_fill_manual(values = hox_palette, drop = FALSE) +
  scale_x_continuous(
    breaks = seq_along(slot_order),
    labels = 15:1,
    expand = expansion(mult = 0)
  ) +
  scale_y_continuous(
    limits = c(0, 2),
    expand = expansion(mult = 0)
  ) +
  labs(
    x = "Hox paralog slot (posterior to anterior)",
    y = NULL,
    caption = paste0(
      "Box position = Hox paralog slot; arrow = transcription direction. ",
      "The ancestral amphioxus cluster is shown under HOXA for layout only."
    )
  ) +
  theme_minimal(base_size = 8) +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(colour = "#D0D0D0", fill = NA, linewidth = 0.25),
    panel.spacing = grid::unit(0.45, "lines"),
    axis.text.x = element_text(size = 6.5),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    strip.background = element_rect(fill = "#ECECEC", colour = "#C8C8C8", linewidth = 0.25),
    strip.text.x = element_text(face = "bold", size = 8),
    strip.text.y.left = element_text(angle = 0, hjust = 1, size = 7.5),
    strip.placement = "outside",
    legend.position = "none",
    plot.caption = element_text(size = 6.5, colour = "#555555", hjust = 0),
    plot.margin = margin(5.5, 8, 5.5, 5.5)
  )
```

The grey amphioxus/HOXB--HOXD cells are structural blanks. The pale-red
zebrafish-b/HOXD cell is different: zebrafish retains HOXAA,
HOXAB, HOXBA, HOXBB, HOXCA, HOXCB, and HOXDA, but not HOXDB. Within a retained
cluster, an empty vertical slot means that the bundled annotation has no
plotted gene at that paralog position. Consult `hox_annotation_gaps.tsv` before
interpreting any such empty slot as evolutionary loss.

The gap audit compares the pinned annotations with the published chicken Hox
inventory ([Liang et al. 2011](https://pmc.ncbi.nlm.nih.gov/articles/PMC3038165/))
and the 43-gene spotted-gar inventory
([Braasch et al. 2016](https://www.nature.com/articles/ng.3526)). These are
source-annotation gaps, not fabricated gene models or automatic loss calls.

```{r hox-annotation-gap-table, echo = FALSE}
gap_groups <- split(
  annotation_gaps,
  interaction(annotation_gaps$species, annotation_gaps$cluster, drop = TRUE)
)
gap_table <- do.call(rbind, lapply(gap_groups, function(rows) {
  display_name <- species$display_name[match(rows$species[[1]], species$species)]
  data.frame(
    Species = display_name,
    Cluster = if (rows$cluster[[1]] == "ancestral") "Ancestral" else paste0("HOX", rows$cluster[[1]]),
    `Source-annotation gap slots` = paste0(
      "Hox",
      paste(sort(unique(rows$hox_number), decreasing = TRUE), collapse = ", Hox")
    ),
    check.names = FALSE
  )
}))
row.names(gap_table) <- NULL
knitr::kable(
  gap_table,
  align = c("l", "l", "l"),
  caption = paste(
    "Literature-expected Hox members without a safe coding model in the",
    "pinned source annotation. Full evidence and candidate-model assessments",
    "remain in hox_annotation_gaps.tsv."
  )
)
```

Two spotted-gar Hox14 positions have different, explicit states rather than
annotation gaps: HoxA14 is a lineage absence, while HoxD14 is a recognizable
pseudogene and is not drawn as a coding gene box. Both are recorded in
`hox_slot_states.tsv`.

The gar HoxA3 and HoxB3 loci also require explicit transcript-level safety
overrides. Their longest Ensembl predictions merge three Hox4-, Hox3-, and
Hox2-like homeodomains, so the builder excludes those unsafe chimeric isoforms
and uses the shorter coherent Hox3 coding isoform at each locus. A separate
HoxC6 model merges HoxC9- and HoxC6-like regions; because no coherent
alternative transcript exists, HoxC6 remains an annotation gap. The merged
candidates stay attached to every affected gap audit row rather than being
silently discarded.

Amphioxus posterior labels deserve special care. They are manual collinear
positions in the BraLan2 cluster, not a claim of strict one-to-one orthology to
every vertebrate posterior paralog group. Hox13 is retained as an empty
annotation slot: the BraLan2 GTF has no corresponding gene model, consistent
with the failure to identify *B. lanceolatum* Hox13 in the expression study of
[Pascual-Anaya et al. 2012](https://pmc.ncbi.nlm.nih.gov/articles/PMC3534614/).

# What does the anchor mean?

For Syn-backed data, `geom_genebox(anchor = ...)` first keeps protein-coding
transcripts with usable CDS, then selects the transcript having the greatest
genomic span for each gene. The bundled curated tables apply that rule after
the explicit gar merged-model safety exclusions described above. Explicit
`start_codon` and `stop_codon` records
define codon-centre anchors. When either record is unavailable, the
corresponding terminal CDS triplet is used as a flagged positional proxy;
consult the anchor-source and fallback columns before interpreting it as a
confirmed initiation or stop codon.

- `anchor = "start"` is the middle nucleotide of an annotated
  translation-initiation codon (ATG in the standard case), or the flagged
  terminal-CDS proxy when that annotation is unavailable.
- `anchor = "end"` is the middle nucleotide of an annotated stop codon, or the
  flagged terminal-CDS proxy when that annotation is unavailable.
- `anchor = "middle"` is their arithmetic genomic-coordinate midpoint. It is
  not the transcript midpoint, UTR midpoint, or midpoint along spliced CDS.

The main matrix uses `middle`. Here the same human HOXA genes are drawn at all
three raw anchors. The x axis is reversed only for presentation, so the panel
still reads Hox13 to Hox1 like the matrix; the displayed numbers remain the
original GRCh38 coordinates. Labels are composed with a separate `geom_text()`
layer because `geom_genebox()` deliberately does not contain text.

```{r human-hoxa-anchors, message = FALSE, warning = FALSE, fig.width = 10.5, fig.height = 4.8, fig.alt = "Three stacked raw-coordinate human HOXA panels place the same fixed-size gene boxes at initiation-codon, genomic coding-midpoint, and stop-codon anchors. The genomic x axis is reversed so HOXA13 is at left and HOXA1 is at right.", fig.cap = "Human HOXA anchor modes on the original GRCh38 coordinate scale. The descending x axis preserves the posterior-to-anterior reading direction; no strip scale is applied."}
human_hoxa <- genes[genes$species == "human" & genes$cluster == "A", ]
anchor_columns <- c(
  "ATG anchor" = "genomic_x_start",
  "Genomic midpoint" = "genomic_x_middle",
  "Stop anchor" = "genomic_x_end"
)
anchor_demo <- do.call(rbind, lapply(names(anchor_columns), function(mode) {
  out <- human_hoxa
  out$x <- out[[anchor_columns[[mode]]]]
  out$anchor_mode <- mode
  out
}))
anchor_demo$anchor_mode <- factor(anchor_demo$anchor_mode, levels = names(anchor_columns))
anchor_demo$track <- as.character(anchor_demo$anchor_mode)
anchor_demo$y <- 1

ggexon() +
  geom_genebox(
    data = anchor_demo,
    aes(fill = slot),
    box_size = 3,
    colour = "grey20",
    linewidth = 0.25,
    show.legend = FALSE
  ) +
  geom_text(
    data = anchor_demo,
    aes(x = x, y = 1.42, label = gene_symbol),
    inherit.aes = FALSE,
    angle = 35,
    hjust = 0,
    size = 2.2,
    check_overlap = TRUE
  ) +
  facet_grid(rows = vars(anchor_mode), switch = "y") +
  scale_fill_manual(values = hox_palette, drop = FALSE) +
  scale_x_reverse(
    labels = function(x) paste0(format(round(x / 1e6, 3), nsmall = 3), " Mb"),
    expand = expansion(mult = c(0.06, 0.06))
  ) +
  scale_y_continuous(limits = c(0.45, 1.7), breaks = NULL, expand = expansion(mult = 0)) +
  labs(x = "GRCh38 coordinate (descending)", y = NULL) +
  theme_minimal(base_size = 8) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "#D0D0D0", fill = NA, linewidth = 0.25),
    strip.background = element_rect(fill = "#ECECEC", colour = "#C8C8C8", linewidth = 0.25),
    strip.text.y.left = element_text(angle = 0, hjust = 1),
    strip.placement = "outside",
    legend.position = "none"
  )
```

```{r human-hoxa-anchor-table, echo = FALSE}
anchor_table <- human_hoxa[order(human_hoxa$hox_number, decreasing = TRUE), ]
anchor_table <- data.frame(
  Gene = anchor_table$gene_symbol,
  `Selected transcript` = anchor_table$transcript_stable_id_version,
  Strand = anchor_table$strand,
  `Transcript span (bp)` = format(anchor_table$transcript_span_bp, big.mark = ","),
  `ATG anchor` = format(anchor_table$genomic_x_start, big.mark = ","),
  Midpoint = format(anchor_table$genomic_x_middle, big.mark = ","),
  `Stop anchor` = format(anchor_table$genomic_x_end, big.mark = ","),
  check.names = FALSE
)
knitr::kable(anchor_table, align = c("l", "l", "c", "r", "r", "r", "r"))
```

# Pinned data and audit trail

The vertebrate annotations are pinned to Ensembl release 116. Amphioxus uses
BraLan2 from Ensembl Metazoa release 63. The builder stores the source URL,
assembly accession, retrieval date, and SHA-256 checksum rather than silently
following a moving current release.

```{r hox-source-table, echo = FALSE}
source_table <- data.frame(
  Species = species$display_name,
  Assembly = species$assembly,
  Source = paste(species$source_database, species$release),
  `Retrieved on` = species$retrieved_on,
  `Source annotation` = sprintf("[%s](%s)", species$source_file, species$source_url),
  check.names = FALSE
)
knitr::kable(source_table, escape = FALSE)
```

The bundled directory contains:

- `hox_genes.tsv`: one selected, plot-ready coding transcript per displayed
  Hox gene, including all three anchors and the selection rule;
- `hox_cds.tsv`: CDS pieces for every selected transcript;
- `hox_clusters.tsv`: the complete 7-by-4 matrix and cell states;
- `hox_annotation_gaps.tsv`: Hox annotations that could not produce a usable
  coding anchor, kept separate from evolutionary absence;
- `hox_expected_complement.tsv`: the source-backed expected functional
  inventory used to distinguish plotted models from annotation gaps;
- `hox_slot_states.tsv`: explicit non-functional states, including absent gar
  HoxA14 and the recognizable but unplotted gar HoxD14 pseudogene;
- `curated_transcript_exclusions.tsv`: the three unsafe merged gar
  transcripts, affected Hox slots, proteins, and exclusion evidence;
- `hox_xref_conflicts.tsv`: the retained source-label disagreement for the
  curated gar HoxA3 translation;
- `manual_hox_mapping.tsv`: explicit stable-ID rescues for unnamed but
  positionally unambiguous Ensembl models;
- `amphioxus_hox_mapping.tsv`: the explicit BraLan2 collinear-position mapping;
- `hox_species.tsv`: releases, assemblies, source URLs, checksums, and retrieval
  dates; and
- `annotations/*.gff3`: compact, original-coordinate records for auditing the
  selected genes and transcripts.

The reproducible builder lives in `data-raw/hox_cluster_expansion`. Its output
is meant to be inspectable: changes in an upstream release should require an
explicit rebuild and review, not silently alter the tutorial.
