---
title: "HOXA Synteny Demo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HOXA Synteny Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9.5,
  fig.height = 7.2,
  dpi = 220
)
```

# Overview

This demo uses a small curated Ensembl release 115 dataset to show HOXA/Hoxa
gene synteny across human, rhesus macaque, mouse, chicken, and green anole. The
dataset is bundled as plot-ready TSV files under `inst/extdata/hoxa_ensembl115`.

The figure uses `geom_genetag()` for HOXA gene intervals and
`geom_synteny_link()` for orthology/synteny ribbons between adjacent species.
The plotted x coordinates are oriented so each cluster reads in a comparable
HOXA13-to-HOXA1 order. Original Ensembl coordinates are retained in the TSV
files.

```{r hoxa-demo, fig.alt = "A ggexon HOXA synteny plot with five horizontal species tracks for human, rhesus macaque, mouse, chicken, and green anole, connected by colored orthology ribbons between adjacent species."}
library(ggexon)

demo_dir <- system.file("extdata", "hoxa_ensembl115", package = "ggexon")
genes <- read.delim(file.path(demo_dir, "hoxa_genes.tsv"), check.names = FALSE)
links <- read.delim(file.path(demo_dir, "hoxa_links.tsv"), check.names = FALSE)
species <- read.delim(file.path(demo_dir, "hoxa_species.tsv"), check.names = FALSE)

track_levels <- as.vector(rbind(
  species$species[-nrow(species)],
  paste0("link_", species$species[-nrow(species)], "_", species$species[-1])
))
track_levels <- c(track_levels, species$species[nrow(species)])

track_labels <- setNames(track_levels, track_levels)
track_labels[species$species] <- paste0(
  species$display_name,
  " (",
  species$source_seqname,
  ")"
)

hox_levels <- paste0("HOXA", c(13, 11, 10, 9, 7, 6, 5, 4, 3, 2, 1))
hox_palette <- setNames(
  grDevices::hcl.colors(length(hox_levels), "Dark 3"),
  hox_levels
)

genes$track <- factor(genes$track, levels = track_levels)
links$track <- factor(links$track, levels = track_levels)
genes$hox_group <- factor(genes$hox_group, levels = hox_levels)
links$hox_group <- factor(links$hox_group, levels = hox_levels)

ggexon() +
  geom_synteny_link(
    data = links,
    mapping = aes(
      tspecies = tspecies,
      tchr = tchr,
      tstart = tstart,
      tend = tend,
      qspecies = qspecies,
      qchr = qchr,
      qstart = qstart,
      qend = qend,
      strand = strand,
      group = group,
      fill = hox_group
    ),
    colour = NA,
    alpha = 0.32,
    inherit.aes = FALSE,
    show.legend = FALSE
  ) +
  geom_genetag(
    data = genes,
    mapping = aes(fill = hox_group),
    exon_height = 0.58,
    arrow_fraction = 0.16,
    show_label = TRUE,
    label_position = "outside",
    label_direction = "top:bottom",
    label_size = 2.15,
    label_link = FALSE,
    label_max_lanes = 2,
    label_panel_width = 175,
    linewidth = 0.18,
    colour = "grey20"
  ) +
  facet_genomics(
    vars(track),
    scales = "free_x",
    ncol = 1,
    labeller = ggplot2::as_labeller(track_labels)
  ) +
  scale_fill_manual(values = hox_palette, drop = FALSE, name = "HOXA group") +
  scale_x_continuous(
    labels = function(x) paste0(round(x / 1000), " kb"),
    expand = ggplot2::expansion(mult = c(0.015, 0.015))
  ) +
  labs(x = "Oriented position in HOXA window", y = NULL) +
  theme_ggexon_track(
    base_size = 8,
    show_x_axis = TRUE,
    show_x_grid = TRUE,
    show_legend = TRUE
  ) +
  theme(
    panel.spacing.y = grid::unit(0.05, "lines"),
    strip.text.y = element_text(angle = 0, hjust = 0, face = "bold", size = 7.5),
    strip.background = element_rect(fill = "grey96", colour = "grey82", linewidth = 0.25),
    legend.position = "bottom",
    legend.key.height = grid::unit(3, "mm"),
    legend.key.width = grid::unit(6, "mm"),
    plot.margin = margin(6, 8, 6, 6)
  )
```

# Provenance

The data-preparation script is stored in `data-raw/hoxa_ensembl115` in the
source repository. Green anole uses the full Ensembl GTF because its HOXA genes
are annotated on scaffold `GL343275.1` and are absent from the chromosome-only
GTF in release 115.
