---
title: "HOXA Synteny Demo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HOXA Synteny Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9.5,
  fig.height = 7.2,
  dpi = 220
)
```

# Overview

This demo uses a small curated Ensembl release 115 dataset to show HOXA/Hoxa
gene synteny across human, rhesus macaque, mouse, chicken, and green anole. It
is designed as a vertebrate example for comparative-genomics users who want to
see conserved gene order, rearranged genomic coordinates, and cross-species
orthology in one figure.

The dataset is bundled as plot-ready TSV files under
`inst/extdata/hoxa_ensembl115`. The figure uses `geom_genetag()` for HOXA gene
intervals and `geom_synteny_link()` for orthology/synteny ribbons between
adjacent species. The plotted x coordinates are oriented so each cluster reads
in a comparable HOXA13-to-HOXA1 order. Original Ensembl coordinates are retained
in the TSV files.

```{r hoxa-demo, message = FALSE, warning = FALSE, fig.alt = "A ggexon HOXA synteny plot with five horizontal species tracks for human, rhesus macaque, mouse, chicken, and green anole, connected by colored orthology ribbons between adjacent species.", fig.cap = "HOXA/Hoxa gene synteny across five vertebrates. Gene intervals are drawn on species tracks and matching HOXA groups are connected by orthology/synteny ribbons."}
library(ggexon)

demo_dir <- system.file("extdata", "hoxa_ensembl115", package = "ggexon")
genes <- read.delim(file.path(demo_dir, "hoxa_genes.tsv"), check.names = FALSE)
links <- read.delim(file.path(demo_dir, "hoxa_links.tsv"), check.names = FALSE)
species <- read.delim(file.path(demo_dir, "hoxa_species.tsv"), check.names = FALSE)
homology <- read.delim(file.path(demo_dir, "hoxa_homology.tsv"), check.names = FALSE)

# Insert one ribbon-only panel between every adjacent pair of species tracks.
species_tracks <- species$species
link_tracks <- paste0("link_", head(species_tracks, -1), "_", tail(species_tracks, -1))
track_levels <- as.vector(rbind(head(species_tracks, -1), link_tracks))
track_levels <- c(track_levels, tail(species_tracks, 1))

track_labels <- setNames(track_levels, track_levels)
track_labels[species_tracks] <- sprintf(
  "%s (%s)",
  species$display_name,
  species$source_seqname
)

hox_levels <- unique(genes$hox_group)
hox_levels <- hox_levels[
  order(as.integer(sub("^HOXA", "", hox_levels)), decreasing = TRUE)
]
hox_palette <- setNames(
  grDevices::hcl.colors(length(hox_levels), "Dark 3"),
  hox_levels
)

genes$track <- factor(genes$track, levels = track_levels)
links$track <- factor(links$track, levels = track_levels)
genes$hox_group <- factor(genes$hox_group, levels = hox_levels)
links$hox_group <- factor(links$hox_group, levels = hox_levels)

ggexon() +
  geom_synteny_link(
    data = links,
    mapping = aes(
      tspecies = tspecies,
      tchr = tchr,
      tstart = tstart,
      tend = tend,
      qspecies = qspecies,
      qchr = qchr,
      qstart = qstart,
      qend = qend,
      strand = strand,
      group = group,
      fill = hox_group
    ),
    colour = NA,
    alpha = 0.32,
    inherit.aes = FALSE,
    show.legend = FALSE
  ) +
  geom_genetag(
    data = genes,
    mapping = aes(fill = hox_group),
    exon_height = 0.58,
    arrow_fraction = 0.16,
    gene_layout = "nested",
    gene_lane_gap = 0.12,
    show_label = TRUE,
    label_position = "outside",
    label_direction = "top:bottom",
    label_size = 2.15,
    label_link = FALSE,
    label_max_lanes = 2,
    label_panel_width = 175,
    linewidth = 0.18,
    colour = "grey20"
  ) +
  facet_genomics(
    vars(track),
    scales = "free_x",
    ncol = 1,
    link_panel_height = 0.32,
    link_axis = "none",
    link_strip = "blank",
    annotation_axis = "bottom",
    strip.position = "left",
    labeller = ggplot2::as_labeller(track_labels)
  ) +
  scale_fill_manual(values = hox_palette, drop = FALSE, name = "HOXA group") +
  scale_x_continuous(
    labels = function(x) paste0(round(x / 1000), " kb"),
    expand = ggplot2::expansion(mult = c(0.015, 0.015))
  ) +
  labs(x = "Oriented position in HOXA window", y = NULL) +
  theme_ggexon_track(
    base_size = 8,
    show_x_axis = TRUE,
    show_x_grid = TRUE,
    show_legend = TRUE
  ) +
  theme(
    panel.spacing.y = grid::unit(0.05, "lines"),
    legend.position = "bottom",
    legend.key.height = grid::unit(3, "mm"),
    legend.key.width = grid::unit(6, "mm"),
    plot.margin = margin(6, 8, 6, 6)
  ) +
  theme_ggexon_side_strips("left", base_size = 7.5)
```

# What the plot shows

Each species has one annotation panel. The narrow panels between species are
link panels, where each ribbon connects the same HOXA group in the two adjacent
species. The figure keeps the original chromosome or scaffold label in the strip
text, but uses oriented plotting coordinates so readers can compare gene order
rather than mentally reversing minus-strand clusters. Link panels are compact
and do not draw their own axes or strip labels, keeping the visual emphasis on
the species annotations while preserving the orthology/synteny ribbons.

The demo uses Ensembl gene-level intervals, so broad gene spans can overlap
when transcript models include long introns or UTRs. For example, chicken
`HOXA10` is nested within the broad `HOXA9` gene span in release 115.
`gene_layout = "nested"` assigns overlapping gene bodies to sublanes so the
annotation structure remains visible instead of hiding one gene behind another.

The input files are deliberately simple:

- `hoxa_genes.tsv` has one row per plotted HOXA/Hoxa gene interval.
- `hoxa_links.tsv` has one row per adjacent-species matched interval.
- `hoxa_species.tsv` records display names, assemblies, source URLs, source
  seqnames, and source notes.
- `hoxa_homology.tsv` maps non-human Ensembl gene IDs to human HOXA reference
  genes for `HomologyAnnotation()`.
- `annotations/*.gff3` contains tiny original-coordinate GFF3 files for building
  a `SynSpecies` object without downloading the full Ensembl GTFs.

The plotting recipe mirrors those tables. First, the species order defines the
annotation tracks. Second, synthetic `link_*` tracks are inserted between
adjacent species so ribbons have their own compact panels. Third, HOXA group
names are ordered by their numeric suffix, which keeps the legend and colors in
the same HOXA13-to-HOXA1 direction as the oriented genomic windows.

# How the link layer runs

`geom_synteny_link()` is a semantic wrapper around `geom_nuclink()`. The wrapper
is useful when the input represents gene-level synteny or orthology intervals
rather than nucleotide alignment fragments.

For an ordinary data frame, the link table supplies target and query intervals:
`tspecies`, `tchr`, `tstart`, `tend`, `qspecies`, `qchr`, `qstart`, `qend`, and
`strand`. Additional columns such as `hox_group` can be mapped to aesthetics
like `fill`.

During plot build, `facet_genomics()` creates the species annotation panels and
the middle link panels. It then attaches the source-panel metadata needed by the
link layer, including `t_panel`, `q_panel`, `target_anchor_y`, and
`query_anchor_y`. `GeomNucLink` converts each interval pair into four polygon
vertices by melting `tstart`, `tend`, `qstart`, and `qend`; target x coordinates
are transformed through the target species panel, and query x coordinates are
transformed through the query species panel. The result is one filled polygon
per matched interval.

This means users can start from a small biological relationship table, keep the
gene annotation and synteny links separate, and let ggexon handle panel layout,
coordinate scaling, and ribbon construction.

# Strip-scale homology view

The same dataset can also be used as a real Syn-backed example. Here the small
GFF3 files keep the original Ensembl coordinates, while `hoxa_homology.tsv`
populates `SynSpecies` homology annotations. `strip_scale_x()` then removes
genomic scale from the gene tracks and uses human as the reference track for
homology-aware gene-block alignment.

```{r hoxa-strip-scale, message = FALSE, warning = FALSE, fig.alt = "A ggexon strip-scale HOXA plot with five species tracks and no ribbons. Gene blocks are aligned by human HOXA reference homology, with x-axis scale hidden.", fig.cap = "Scale-stripped HOXA/Hoxa gene blocks aligned by human reference homology. The plot uses original-coordinate GFF3 inputs and stored HomologyAnnotation mappings, but no synteny ribbons."}
annotation_dir <- file.path(demo_dir, "annotations")

hoxa_syn <- SynSpecies(name = "HOXA Ensembl 115")
for (species_id in species$species) {
  hoxa_syn <- add_individual(
    hoxa_syn,
    SynIndividual(
      annotation_file = file.path(annotation_dir, paste0(species_id, ".gff3")),
      genome_file = genome_waiver(),
      id = species_id,
      annotation_format = "gff"
    )
  )
}

for (query_species in unique(homology$query_species)) {
  rows <- homology[homology$query_species == query_species, , drop = FALSE]
  hoxa_syn <- add_homology_annotation(
    hoxa_syn,
    HomologyAnnotation(
      name = paste0(query_species, "_to_human"),
      reference_species = "human",
      query_species = query_species,
      homology_table = rows[
        c("query_gene", "reference_gene", "hox_group", "reference_gene_id")
      ]
    )
  )
}

syn_track_labels <- setNames(
  sprintf("%s (%s)", species$display_name, species$source_seqname),
  species$species
)

strip_plot <- ggexon(hoxa_syn)
for (species_id in species$species) {
  strip_plot <- strip_plot +
    geom_genetag(
      species = species_id,
      feature_type = "gene",
      mapping = aes(fill = reference_gene),
      exon_height = 0.58,
      arrow_fraction = 0.16,
      gene_layout = "nested",
      gene_lane_gap = 0.12,
      show_label = TRUE,
      label_filter = "homology_hit",
      label_position = "outside",
      label_direction = "top:bottom",
      label_size = 2.15,
      label_link = FALSE,
      label_max_lanes = 2,
      label_panel_width = 175,
      linewidth = 0.18,
      colour = "grey20"
    )
}

strip_plot +
  strip_scale_x(
    reference_track = "human",
    gene_order = "reference",
    guide = "none"
  ) +
  facet_genomics(
    vars(track),
    scales = "free_x",
    ncol = 1,
    annotation_axis = "bottom",
    strip.position = "left",
    labeller = ggplot2::as_labeller(syn_track_labels)
  ) +
  scale_fill_manual(values = hox_palette, drop = FALSE, name = "Human HOXA reference") +
  scale_x_continuous(expand = ggplot2::expansion(mult = c(0.015, 0.015))) +
  labs(x = NULL, y = NULL) +
  theme_ggexon_track(
    base_size = 8,
    show_x_axis = FALSE,
    show_x_grid = FALSE,
    show_legend = TRUE
  ) +
  theme(
    panel.spacing.y = grid::unit(0.05, "lines"),
    legend.position = "bottom",
    legend.key.height = grid::unit(3, "mm"),
    legend.key.width = grid::unit(6, "mm"),
    plot.margin = margin(6, 8, 6, 6)
  ) +
  theme_ggexon_side_strips("left", base_size = 7.5)
```

# Provenance

The data-preparation script is stored in `data-raw/hoxa_ensembl115` in the
source repository. It downloads the Ensembl release 115 GTF files, extracts gene
features whose normalized gene names match `HOXA[0-9]+`, orients plotting
coordinates for the ribbon example, writes original-coordinate GFF3 files for
Syn-backed examples, and exports the homology table used above.

Green anole uses the full Ensembl GTF because its HOXA genes are annotated on
scaffold `GL343275.1` and are absent from the chromosome-only GTF in release 115.
